

This R code accompanies the manuscript "Shifts in predator behavior following climate induced disturbance on coral reefs" by Rotjan & Ray et al.
the dataset can be downloaded from Figshare repository here: https://figshare.com/s/71a3e9133a2de60e96e3

Load the dataset
```{r}
# the dataset can be downloaded from Figshare repository here: https://figshare.com/s/71a3e9133a2de60e96e3

library(readr)
bleaching_coralivory_data_take_2 <- read_csv("C:/Users/nray1/Dropbox/Bleaching Corralivory/Stats & Good Data Sheets/bleaching_coralivory_data - take 2.csv")
View(bleaching_coralivory_data_take_2)

bleach = bleaching_coralivory_data_take_2
```

Load required libraries
```{r}
library(ggplot2)
library(lme4)
library(pscl)
library(dplyr)
library(patchwork)
library(emmeans)
```


Do parrotfish selectively avoid grazing bleached Orbicella corals?
```{r}
#load the dataset - sheet "grazing_observations_2019"
library(readxl)
bleaching_coralivory_data_proc_b_submission <- read_excel("C:/Users/nray1/Desktop/bleaching_coralivory_data - proc_b submission.xlsx", 
    sheet = "grazing_observations_2019")
View(bleaching_coralivory_data_proc_b_submission)

#rename it "bleach"
bleach = bleaching_coralivory_data_proc_b_submission

#compare bites on grazed v. ungrazed corals
#histogram first - do grazed and ungrazed incidences
#make a new vector:
bleached_scars = as.data.frame(bleach$scars_in_bleach)
bleached_label = as.vector("Bleached")
bleached_scars$condition = bleached_label
names(bleached_scars)[names(bleached_scars) == "bleach$scars_in_bleach"] <- "number"
bleached_scars

unbleached_scars = as.data.frame(bleach$num_scars_unbleached)
unbleached_label = as.vector("Unbleached")
unbleached_scars$condition = unbleached_label
names(unbleached_scars)[names(unbleached_scars) == "bleach$num_scars_unbleached"] <- "number"
unbleached_scars

type_hist_data = rbind(bleached_scars, unbleached_scars)

#Try running a hurdle model to compare bites on blecahed/unbleached sections:
bite.mod.hurdle <- hurdle(number ~ condition, data = type_hist_data)
summary(bite.mod.hurdle)

#make a figure of the average number of bites on bitten corals
##make a new dataframe without zeros
no_zeros <- filter(type_hist_data, number > 0)
no_zeros

no_zeros$condition = factor(no_zeros$condition, levels = c("Unbleached", "Bleached"))
max(no_zeros$number)

bites_comp = ggplot(no_zeros, aes(y = number, x = condition, fill = condition))+
  geom_boxplot(outlier.shape = NA)+
  theme_classic()+
  geom_jitter(width = 0.2)+
  theme(axis.text.x =  element_text(size = 10,colour = "black"),
         axis.text.y =  element_text(size = 10,colour = "black"),
         axis.title.y = element_text(size = 10, colour = "black"),
         axis.title.x = element_blank())+
  scale_fill_manual(values = c("#745900", "#f6f3c1"))+
  labs(y = expression(paste("Scars on Grazed Sections")), 
        x = "", color = "white")+
  guides(fill = FALSE)+
  scale_x_discrete(limits = c("Unbleached", "Bleached"))+
  annotate("text", label = "p < 0.001", x = 2, y = 200, size = 3.528)+
  scale_y_continuous(limits = c(0,250))

bites_comp

#################
max(na.omit(type_hist_data$number))
###figure out how to bin the data on the next two lines...

zeros = filter(type_hist_data, number == 0)
zeros

bleach_zero = subset(zeros, condition == "Bleached")
summary(bleach_zero)

unbleach_zero = subset(zeros, condition == "Unbleached")
summary(unbleach_zero)

binomial_bites <- data.frame("Condition" = c("Bleached","Unbleached", "Bleached","Unbleached"),
                "X-axis label" = c("0", "0", "1+", "1+"),
                "Number" = c(143,113,13,58))
binomial_bites$condition_number = paste(binomial_bites$Condition, binomial_bites$X.axis.label)
binomial_bites

##unstack the histogram and turn into just 0 and >=1 (with bars for bleach and unbleach)
bitehist = ggplot(binomial_bites, aes(condition_number, Number, 
                                      fill = condition_number), stat = "identity") +
  geom_col(colour = "black")+
  theme_classic()+
  theme(axis.text.x =  element_text(size = 10,colour = "black"),
         axis.text.y =  element_text(size = 10,colour = "black"),
         axis.title.y = element_text(size = 10, colour = "black"),
         axis.title.x = element_blank())+
  scale_x_discrete(limits = c("Unbleached 0", "Bleached 0", "Unbleached 1+", "Bleached 1+"))+
  scale_fill_manual(values = c("#f6f3c1", "#f6f3c1","#745900", "#745900"))+
  labs(y = expression(paste("Number of Coral Sections")))+
  guides(fill = FALSE)+
  scale_y_continuous(limits = c(0,160))+
  annotate("text", label = "p < 0.001", x = 4, y = 128, size = 3.528)+
  annotate("text", label = "n = 113", x = 1, y = 7, size = 3.528)+
  annotate("text", label = "n = 143", x = 2, y = 7, size = 3.528)+
  annotate("text", label = "n = 58", x = 3, y = 7, size = 3.528)+
  annotate("text", label = "n = 13", x = 4, y = 7, size = 3.528)

bitehist

bitehist_relab = 
  bitehist + scale_x_discrete(labels=c("Unbleached 0" = "0 Scars",
                                       "Bleached 0" = "0 Scars",
                                       "Unbleached 1+" = "1+ Scars", 
                                       "Bleached 1+" = "1+ Scars"),
                              limits = c("Unbleached 0", "Bleached 0", 
                                         "Unbleached 1+", "Bleached 1+"))
bitehist_relab

###now put the two figures together (this becomes Figure 3 in the manuscript)
bite_bleach_plot = bitehist_relab + bites_comp +
  plot_layout(ncol = 2, widths = c(1,1))+
  plot_annotation(tag_levels = "A", tag_suffix = ".")& 
  theme(plot.tag = element_text(size = 10))
bite_bleach_plot

ggsave(filename = "bleach_bites_hist_box.png", plot = bite_bleach_plot, width = 190, height = 100, units = "mm")
```

Does the severity of bleaching relate to the number of bite scars?
```{r}
#here, use the sheet "bites_bleach_score"
library(readxl)
bleaching_coralivory_data_proc_b_submission <- read_excel("C:/Users/nray1/Desktop/bleaching_coralivory_data - proc_b submission.xlsx", 
    sheet = "bites_bleach_score")
View(bleaching_coralivory_data_proc_b_submission)

#rename the datasheet
grazing_bleach_score = bleaching_coralivory_data_proc_b_submission

#estimate a lm
#first make the values into a factor
grazing_bleach_score$bleaching_score = as.factor(grazing_bleach_score$bleaching_score)
model1 = lm(bites_in_section ~ bleaching_score, data = grazing_bleach_score)
summary(model1)

#supplemental table 1
lsmeans(model1, pairwise~bleaching_score)

#manuscript figure 4
graze_score = ggplot(grazing_bleach_score, aes(y = bites_in_section, x = bleaching_score, fill = bleaching_score))+
  geom_boxplot(outlier.shape = NA)+
  theme_classic()+
  geom_jitter(width = 0.2)+
  theme(axis.text.x =  element_text(size = 10,colour = "black"),
         axis.text.y =  element_text(size = 10,colour = "black"),
         axis.title.y = element_text(size = 10, colour = "black"),
         axis.title.x = element_text(size = 10, colour = "black"))+
  scale_fill_manual(values = c("#f6f3c1", "#f5e9bf", "#eed688", "#cfaf48", "#9c801f","#745900"))+
  labs(y = expression(paste("Scars on Grazed Sections")), 
        x = "Coral Bleaching Score")+
  guides(fill = FALSE)+
  scale_y_continuous(limits = c(0,250))+
  scale_x_discrete(limits = c(1,2,3,4,5,6))+
  annotate("text", x = 1, y = 10, label = "n.s.", size = 3.528, fontface = 'italic')+
  annotate("text", x = 1, y = 250, label = "Bleached", size = 3.528, hjust = 0)+
  annotate("text", x = 4, y = 250, label = "Unbleached", size = 3.528, hjust = 0)+
  geom_hline(yintercept = 0)+
  geom_vline(xintercept = 3.5, linetype = "dashed")
graze_score

ggsave(filename="bleach_score_bites.png", plot=graze_score, width = 100, height = 100, units = "mm")

```


Does coral surface area relate to number of bite marks or bleaching?
```{r}
#use the "bleach" dataset loaded previously

#first, calculate the surface area using Knud Thomsen's formula (http://mathcentral.uregina.ca/QQ/database/QQ.09.07/h/paritosh1.html)
p = 1.6075
bleach$surface_area = 4*pi*((((bleach$length_cm^p*bleach$height_cm^p)+(bleach$length_cm^p*bleach$width_cm^p)+(bleach$width_cm^p*bleach$height_cm^p))/3)^(1/p))
bleach$surface_area
#divide by two, since only one side of the eliptical dome is available for grazing/bleaching
bleach$surface_area = bleach$surface_area/2
#then convert to m2 by dividing by 10000
bleach$surface_area_m2 = bleach$surface_area/10000

max(bleach$surface_area_m2)
min(bleach$surface_area_m2)

#test for relationship between grazing and area:
#make a column of total scars per coral
bleach$total_scars = bleach$num_scars_unbleached + bleach$scars_in_bleach
scar_area = lm(bleach$total_scars ~ bleach$surface_area_m2)
summary(scar_area)
plot(bleach$total_scars ~ bleach$surface_area_m2)
#p < 0.01, but R2 = 0.10

#look at number of scars in bleach and unbleach as a function of surface area
unbleach_scar_area = (lm(bleach$num_scars_unbleached ~ bleach$surface_area_m2))
summary(unbleach_scar_area)
#only describes 6% of variance for unbleached
bleach_scar_area = (lm(bleach$scars_in_bleach ~ bleach$surface_area_m2))
summary(bleach_scar_area)
#nothing for bleached
#pooling the total number of scars is best

#does size predict bleaching susceptibility?
summary(lm(bleach$percent_bleached ~ bleach$surface_area_m2))
plot(bleach$percent_bleached ~ bleach$surface_area_m2)
#no relationship
#p = 0.667 r2 = 0

#make a stacked plot
size_scars = ggplot(data = bleach, aes(x = surface_area_m2, y = total_scars))+
  geom_hline(yintercept = 0)+
  geom_smooth(method = "lm", colour = "black", se = TRUE)+
  geom_point(colour = "#745900")+
  theme_classic()+
  theme(axis.text.x =  element_blank(),
         axis.text.y =  element_text(size = 10,colour = "black"),
         axis.title.y = element_text(size = 10, colour = "black"),
         axis.title.x = element_blank())+
  labs(y = expression(paste("Number of Scars")))+
  scale_y_continuous(limits = c(0, 250),breaks=c(0, 50, 100, 150, 200, 250))+
  guides(fill = FALSE)+
  annotate("text", label = "paste(italic(R)^2, \" = 0.10\")", x = 6, y = 250, size = 3.528, parse = TRUE, hjust = 0)+
  annotate("text", x = 6, y = 220, label = "p < 0.001", size = 3.528, hjust = 0)
size_scars

size_bleaching = ggplot(data = bleach, aes(x = surface_area_m2, y = percent_bleached))+
  geom_hline(yintercept = 0)+
  #geom_smooth(method = "lm", colour = "black", se = FALSE)+
  geom_point(colour = "#745900")+
  theme_classic()+
  theme(axis.text.x =  element_text(size = 10,colour = "black"),
         axis.text.y =  element_text(size = 10,colour = "black"),
         axis.title.y = element_text(size = 10, colour = "black"),
         axis.title.x = element_text(size = 10,colour = "black"))+
  labs(y = expression(paste("% of Colony Bleached")))+
  labs(x = expression(paste("Colony Surface Area (m"^2*")")))+
  scale_y_continuous(limits = c(0, 100),breaks=c(0, 20, 40, 60, 80, 100))+
  guides(fill = FALSE)+
  annotate("text", label = "paste(italic(R)^2, \" = 0.00\")", x = 6, y = 100, size = 3.528, parse = TRUE, hjust = 0)+
  annotate("text", x = 6, y = 88, label = "p = 0.667", size = 3.528, hjust = 0)
size_bleaching

size_relation_plot = size_scars + size_bleaching +
  plot_layout(ncol = 1)+
  plot_annotation(tag_levels = "A", tag_suffix = ".")& 
  theme(plot.tag = element_text(size = 10))
size_relation_plot

#figure 5
ggsave(filename = "coral_szie_relations.png", plot = size_relation_plot, width = 100, height = 150, units = "mm")
```


Does corallivory influence bleaching susceptibility?
```{r}
#load the "before_after_2018_2019" sheet
library(readxl)
bleaching_coralivory_data_proc_b_submission <- read_excel("C:/Users/nray1/Desktop/bleaching_coralivory_data - proc_b submission.xlsx", 
    sheet = "before_after_2018_2019")
View(bleaching_coralivory_data_proc_b_submission)

grazing_mortality = bleaching_coralivory_data_proc_b_submission

#use a lm to compare grazing intensity with % bleaching:
summary(lm(perc_bleach_2019 ~ bites_2018, data = grazing_mortality))
plot(grazing_mortality$perc_bleach_2019 ~ grazing_mortality$bites_2018)

#try using a polynomial fit instead
summary(lm(perc_bleach_2019 ~ log(bites_2018+1), data = grazing_mortality))

graze_bleach = ggplot(data = grazing_mortality, aes(x = bites_2018, y = perc_bleach_2019))+
  geom_hline(yintercept = 0)+
  geom_smooth(method = lm, formula = "y ~ log(x+1)", se = TRUE, colour = "black")+
  geom_point(colour = "#745900")+
  theme_classic()+
  theme(axis.text.x =  element_text(size = 10,colour = "black"),
         axis.text.y =  element_text(size = 10,colour = "black"),
         axis.title.y = element_text(size = 10, colour = "black"),
         axis.title.x = element_text(size = 10,colour = "black"))+
  labs(y = expression(paste("% Bleached in 2019")))+
  labs(x = expression(paste("Number of Bite Scars in 2018")))+
  scale_y_continuous(limits = c(0, 100),breaks=c(0, 20, 40, 60, 80, 100))+
  guides(fill = FALSE)+
  annotate("text", label = "paste(italic(R)^2, \" = 0.19\")", x = 400, y = 50, size = 3.528, parse = TRUE, hjust = 0)+
  annotate("text", x = 400, y = 39.09, label = "p < 0.001", size = 3.528, hjust = 0)
graze_bleach

#figure 6
ggsave(filename = "bleaching_after_bites.png", plot = graze_bleach, width = 100, height = 100, units = "mm")
```


and compare coral %C and %N - this is Table 2
```{r}
#load the sheet labeled "c_n_content"
library(readxl)
bleaching_coralivory_data_proc_b_submission <- read_excel("C:/Users/nray1/Desktop/bleaching_coralivory_data - proc_b submission.xlsx", 
    sheet = "c_n_content")
View(bleaching_coralivory_data_proc_b_submission)

c_n_content = bleaching_coralivory_data_proc_b_submission

c_n_content = rename(c_n_content, site = `Collection Site`)
c_n_content = rename(c_n_content, state = `Symbiotic State`)
c_n_content$site = as.factor(c_n_content$site)
c_n_content$state = as.factor(c_n_content$state)

perc_c_test = lmer(perc_C ~ state + (1|site), data = c_n_content)
summary(perc_c_test)
lsmeans(perc_c_test, pairwise~state)

perc_N_test = lmer(perc_N ~ state + (1|site), data = c_n_content)
summary(perc_N_test)
lsmeans(perc_N_test, pairwise~state)

d13c_test = lmer(d13c ~ state + (1|site), data = c_n_content)
summary(d13c_test)
lsmeans(d13c_test, pairwise~state)

d15n_test = lmer(d15n ~ state + (1|site), data = c_n_content)
summary(d15n_test)
lsmeans(d15n_test, pairwise~state)
```





